D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-01-31 ootavacuous.tex
Update IPC plots to x86 server
Also commit a few PDFs to avoid relying on inkscape converting from .svg
to .pdf, which has odd compatibility issues in recent releases.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This example could make lines~3 and~27 be \co{memory_order_release} and
lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
a pointer to \co{r} before the lifetime of \co{r} has started.
Note that \co{std::unreachable()} does not suffice because the CPU
can also reorder relaxed stores.
Note also that load-store ordering is not sufficient to prevent a
pointer to \co{r} from being stored to \co{b} prior to the start of
\co{r}'s lifetime because there is no atomic load between the start
of that lifetime and the relaxed store to \co{b}.

@@

Finally, there might be a desire for hard evidence that coe and fre
links really can go backwards in time.
We provide this evidence on x86 to demonstrate that these effects
are not confined to weakly ordered systems.
This system is a laptop with Intel(R) Core(TM) i9-10885H CPU @ 2.40GHz,
having a single socket with 8~cores, each having a pair of hardware
threads.
The code generating this data may be found in the \path{CodeSamples/cpu}
directory of the
perfbook~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
	\co{git clone git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}

\begin{figure}[tb]
\begin{center}
\resizebox{3in}{!}{\includegraphics{coe-sh-out}}
\caption{On x86, coe Links Are Atemporal}
\label{fig:On x86, coe Links Are Atemporal}
\end{center}
\end{figure}

Figure~\ref{fig:On x86, coe Links Are Atemporal}
shows the results of 15 of the laptop's hardware threads simultaneously
storing their identifying integer (ranging from 0 to 14) to a shared
variable, then repeatedly doing timestamped reads from that variable.
The x-axis displays time in TSC periods, each of which is about 0.417~ns
in duration.
The y-axis shows the number of distinct opinions that the hardware
threads have as a function of time, a number that is frequently rather
larger than one.
The backwards-in-time coe link from Thread~1 to Thread~0 in
Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}
is therefore entirely plausible.

@@

However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
shows that the order of reads of each value at each thread are
consistent with a number of global orders, as required.
Figure~\ref{fig:On Power5, coe Links Are Also Partially Ordered}
shows a rather more elaborate partial order from a Power5
system~\cite{McKenney2018ParallelProgramming-2018-12-08a}.

@@

Figure~\ref{fig:On x86, fre Links Are Atemporal}
plots a histogram of the elapsed time from the last load to return the
old value to the store that provided the new value.
Note well that most of the data falls into negative time, that is to say,
the last load of the old value executes about 50 timestamp periods (about
20~nanoseconds) \emph{after} the store that overwrote that same old value.
That is, fre links can and do go backwards in time.
This corresponds to the fre link between Thread~3 and Thread~2 depicted in
Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}.

>>>>

This example could make lines~3 and~27 be \co{memory_order_release} and
lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
a pointer to \co{r} before the lifetime of \co{r} has started.
Alternatively, threads could be spawned only after \co{r}'s lifetime
starts and joined before it ends.
Placing \co{r} in static global storage would avoid lifetime considerations
totally.
And there are of course any number of other synchronization schemes
that might be used to ensure that pointers are in use only while the
pointed-to objects are alive.

Note that \co{std::unreachable()} does not suffice because the CPU
can also reorder relaxed stores.
Note also that load-store ordering is not sufficient to prevent a
pointer to \co{r} from being stored to \co{b} prior to the start of
\co{r}'s lifetime because there is no atomic load between the start
of that lifetime and the relaxed store to \co{b}.

@@

Finally, there might be a desire for hard evidence that coe and fre
links really can go backwards in time.
We provide this evidence on x86 to demonstrate that these effects
are not confined to weakly ordered systems.
This system is a dual-socket system with
Intel(R) Xeon(R) Gold 6138 CPU @ 2.00GHz,
each socket with 20~cores, each having a pair of hardware
threads, for a grand total of 80 hardware threads.
The code generating this data may be found in the \path{CodeSamples/cpu}
directory of the
perfbook~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
	\co{git clone git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}

\begin{figure}[tb]
\begin{center}
\resizebox{4in}{!}{\includegraphics{coe-nvals}}
\caption{On x86, Reasonable CPUs Can Disagree}
\label{fig:On x86; Reasonable CPUs Can Disagree}
\end{center}
\end{figure}

Figure~\ref{fig:On x86; Reasonable CPUs Can Disagree}
shows the results of 79 of the systems's hardware threads simultaneously
storing their identifying integer (ranging from 0 to 78) to a shared
variable, then repeatedly doing timestamped reads from that variable.
The x-axis displays time in TSC periods, each of which is about 0.5~ns
in duration.
The y-axis shows the number of distinct opinions that the hardware
threads have as a function of time, a number that is frequently rather
larger than one.
The backwards-in-time coe link from Thread~1 to Thread~0 in
Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}
is therefore entirely plausible.

\begin{figure}[tb]
\begin{center}
\resizebox{4in}{!}{\includegraphics{coe-sh-out}}
\caption{On x86, coe Links Are Atemporal}
\label{fig:On x86; coe Links Are Atemporal}
\end{center}
\end{figure}

Not only is this plausible, but as you can see in
Figure~\ref{fig:On x86; coe Links Are Atemporal},
it actually happens, and frequently.
The store-to-store latency is measured from the end of the ``winning''
store to the beginning of the non-winning store that started latest.
As you can see, most of the measurements show that the winning store
was not the last store, even on x86.

@@

@@@ However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
shows that the order of reads of each value at each thread are
consistent with a number of global orders, as required.
Figure~\ref{fig:On Power5, coe Links Are Also Partially Ordered}
shows a rather more elaborate partial order from a Power5
system~\cite{McKenney2018ParallelProgramming-2018-12-08a}.

@@

Figure~\ref{fig:On x86, fre Links Are Atemporal}
plots a histogram of the elapsed time from the beginning of load to
return the old value to the end of the store that provided the new value.
Note well that most of the data falls into negative time, that is to
say, most frequently, the last load of the old value executes about 60
timestamp periods (about 30~nanoseconds) \emph{after} the store that
overwrote that same old value.
That is, fre links can and do go backwards in time.
This corresponds to the fre link between Thread~3 and Thread~2 depicted in
Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}.
